\documentclass{standalone}
\usepackage{chez}

\begin{document}
\section{Decidability of \texorpdfstring{\(A_{\textsf{TM}}\)}{A\_TM} and Friends}
\begin{theorem}
	\(A_{\textsf{TM}} = \set{\angles{M, w} \mid \text{$M$ is a \textsf{TM} accepting $w$}}\) is not decidable.
\end{theorem}
\begin{proof}
	Suppose for the sake of contradiction that there exists a \textsf{TM} \(H\) that decides \(A_{\textsf{TM}}\). Let's construct another \textsf{TM} \(D\) that does the following:
	\begin{enumerate}[start=0]
		\item On input \(\angles{M}\), where \(M\) is a \textsf{TM}:
		\item Run \(H\) on \(\angles{M, \angles{M}}\).
		\item If \(H\) accepts then \textsc{reject}, else \textsc{accept}.
	\end{enumerate}
	
	Intuitively, \(D\) accepts \(\angles{M}\) if and only if \(M\) on \(\angles M\) does not accept.

	Now consider running \(D\) on \(\angles{D}\). Then \(D\) accepts if and only if \(D\) on \(\angles{D}\) does not accept. This is a contradiction.
\end{proof}

The motivation comes from Cantor's diagonalization argument. In particular, consider the table:
\[
	\begin{array}{r | c c c c c} % chktex 44
		    & \angles{M_1} & \angles{M_2} & \angles{M_3} & \cdots & \angles{D} \\ \hline % chktex 44
		M_1 & \textsc{acc} & \textsc{rej} & \textsc{acc} & \cdots & \\
		M_2 & \textsc{acc} & \textsc{acc} & \textsc{acc} & \cdots & \\
		M_3 & \textsc{rej} & \textsc{rej} & \textsc{rej} & \cdots & \\
		\vdots\,\, & \vdots & \vdots & \vdots & \ddots & \vdots \\
		D & \textsc{rej} & \textsc{rej} & \textsc{acc} & \cdots & ? \\
	\end{array}
\]
The \textsf{TM} \(D\) gives the opposite of the entries on the main diagonal, which yields a problem when we get to the row for \(D\).

\begin{note}
	Sometimes we \emph{do} input programs into themselves! For example, compilers.
\end{note}

\begin{theorem}
	 If \(A\) and \(\ol{A}\) are both Turing recognizable, then \(A\) is decidable.
\end{theorem}
\begin{proof}
	Let \(R\) be a recognizer for \(A\) and \(S\) be a recognizer for \(\ol A\). Then, consider the following \textsf{TM}:
	\begin{enumerate}[start=0]
		\item On input string \(w\):
		\item Start running \(R\) and \(S\) on \(w\).
		\item Let \(R\) and \(S\) alternate taking steps until one accepts.
		\item If \(R\) accepts, then \textsc{accept}, and if \(S\) accepts, then \textsc{reject}.
	\end{enumerate}
	Note that this is a decider because one of \(R\) and \(S\) must accept, halting the \textsf{TM}.
\end{proof}

\begin{corollary}
	\(\ol{A_{\textsf{TM}}}\) is not Turing recognizable.
\end{corollary}


\end{document}

